Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)

### 2020-07-12 manuscript/manuscript.Rmd
manuscript v0.9

---
title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
bibliography: references.bib
csl: elsevier-vancouver.csl
output:
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
  word_document:
    toc: yes
    reference_docx: template.docx
link-citations: yes
always_allow_html: yes
---

Peter Dutey Magni,^{1} Haydn Williams,^{2} Arnoupe Jhass,^{1} Greta Rait, Harry Hemingway,^{134} Andrew Hayward,^{1} Laura Shallcross^{1}

1 Institute of Health Informatics, University College London, London, UK
2 Four Seasons Health Care Group
3 Health Data Research UK, University College London, London, UK
4 University College London Hospitals NIHR Biomedical Research Centre, London, UK
5 Institute of Epidemiology & Health Care, University College London, London, UK

@@

incidence_CT_residents <- table_nums(
  name = "incidence_CT_residents",
  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
supp_epi_CT_oubreak_only <- table_nums(
  name = "incidence_CT_staff",
  caption =  paste0("Incidence proportions and rates of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_resident_descriptives <- table_nums(
  name = "tab_resident_descriptives",
  caption = paste0("Characteristics of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CFR_age_sex <- table_nums(
  name = "tab_CFR_age_sex",
  caption =  paste0("All-cause case-fatality rates by age and sex among private residents (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  "; ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CoxPH_confirmed <- table_nums(
  name = "tab_CoxPH_confirmed",
  caption =  paste0("Risk factors for confirmed infection in private residents: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  ")"))
tab_CoxPH_death <- table_nums(
  name = "tab_CoxPH_death",
  caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_attributable_deaths <- table_nums(
  name = "tab_attributable_deaths",
  caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))

@@

**Background** Covid-19 has led to a 45% increase in mortality in care home residents in England and Wales, but the proportion of deaths directly attributable to Covid-19 is uncertain due to limited testing for infection, and lack of syndromic surveillance.

**Methods:**  Retrospective cohort study using individual-level electronic records from residents and daily care home counts of infection for `r formatbm(desc_residents$CT_total_residents)` residents and `r formatbm(desc_staff$total_staff)` staff across `r dplyr::n_distinct(residents$home_code)` care homes from the Four Seasons Health Care Group (FSHCG, `r format(date_min, "%e %B %Y")`-`r format(date_max, "%e %B %Y")`). We estimated the incidence of symptoms, confirmed infections, and deaths in residents and staff and modelled risk factors for infection and mortality in residents using Cox proportional hazards.

**Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had confirmed infections. The incidence of confirmed infection in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days.
`r desc_residents$CT_zero_case`/`r desc_residents$total_homes` homes (`r tbrounding( desc_residents$CT_zero_case/desc_residents$total_homes*100)`%) had neither confirmed infections nor Covid-related deaths.

Among `r formatbm(desc_residents$DAT_total_residents)` private residents, there were `r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]), `r desc_residents$DAT_CFR$y` of which occurred in residents with confirmed infection, equivalent to an all-cause case-fatality rate of `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]. `r tbrounding(((desc_residents$DAT_CFR_RR - 1)/desc_residents$DAT_CFR_RR)*100)`% [`r tbrounding(((desc_residents$DAT_CFR_RR_lower - 1)/desc_residents$DAT_CFR_RR_lower)*100)`%; `r tbrounding(((desc_residents$DAT_CFR_RR_upper - 1)/desc_residents$DAT_CFR_RR_upper)*100)`%] of these deaths were attributable to SARS-CoV-2. Symptomatic and asymptomatic infection were independent risk factors for mortality. All-cause mortality for uninfected residents was 93% higher in care homes with outbreaks compared to those without outbreaks (adjusted HR = 1·9 [1·7; 2·2]).

**Interpretation:** Covid-19 infection directly increases residents’ risk of death, but in outbreak care homes mortality is also elevated in individuals with no record of infection compared to homes without outbreaks. Preparations for future waves of the pandemic should prioritise regular testing in care homes to address case under-ascertainment, and consider the indirect effect of Covid-19 on mortality.

**Funding:** This work was supported by the Economic and Social Research Council [grant reference ES/V003887/1].

@@

Care home residents are particularly vulnerable to Covid-19 infection due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and this is compounded by frequent exposure to infection through close contact with infected staff members, other residents and contaminated surfaces in the care home setting. In the UK, an estimated 420,000 residents lived in approximately 11,000 homes for the elderly.[@cqcdata2020]

At the peak of the pandemic the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020]  Data from the Office for National Statistics suggest that two-thirds of the 29,393 excess deaths recorded in care home residents between 28th December 2019 and 12th June 2020 (calculated relative to equivalent period in the previous year) involved Covid-19.[@onsDR2020] However, estimating the proportion of care home deaths that can be directly attributed to Covid-19 based on death certifications is challenging because many care home residents with suspected infection were not tested, particularly at the start of the pandemic. Understanding the ratio of direct to indirect deaths in care home residents will be critical to inform public health measures to mitigate the impact of the anticipated second wave of the pandemic, since lockdown measures[@Hale2020] that prevent infection may have a negative impact on residents by disrupting usual care.

Disentangling direct and indirect deaths from Covid-19 requires accurate information on the incidence and prevalence of infection among staff and residents linked to individual-level clinical outcomes. There is no syndromic surveillance for infection in care homes in England, and universal testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established in care homes in England until 11th May 2020, when the government launched the whole care home testing programme.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England's (PHE) outbreak investigations with a maximum of five tests per care home. Consequently national estimates of incidence and prevalence based on test results obtained prior to 11th May will substantially underestimate the burden of infection in care home residents and staff.

Outbreak investigations provide estimates of the proportion of residents and staff who become infected during an outbreak and case-fatality rates,[@McMichael2020; @Dora2020] but they provide limited insight into the relationship between infection and mortality across the entire pandemic period, and cannot be generalised to care homes that have not reported any outbreaks. A living systematic review[@Salcher-Konrad2020] of outbreak investigations found incidence proportions ranging between 0% and 72% among residents and between 2% and 64% among staff of long-term care facilities. Mortality rates in residents with confirmed Covid-19 infection ranged from 0 to 10%, with case fatality rates between 0 and 34%, however this review did not include estimates of indirect mortality.

We used administrative data from the Four Seasons Health Care Group (FSHCG), one of the UK’s largest independent provider of long-term and respite residential and nursing care, to establish syndromic surveillance for Covid-19 across `r formatbm(dplyr::n_distinct(residents$home_code))` FSHCG homes. We used this system to estimate the incidence and prevalence of suspected and confirmed Covid-19 across a national network of care homes, and to investigate the fraction of all-cause deaths attributable to confirmed infections.

@@

In 2020, the FSHCG included 9,568 residential or nursing care beds for elderly residents, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 3). Approximately 9% of these beds were contracted to local authorities, with the remainder for privately funded residents.

FSHCG collects administrative data on their private residents through a variety of electronic systems that are primarily used for billing purposes. This includes individual-level data on the dates of care home entry and exit for residents, demographic information, and reports of infection through an incidents management system known as “Datix”. Such records do not include residents occupying beds contracted by local authorities, but home-level aggregate information was nevertheless available on all residents (`r fig_nums("fig_study_diagram", display = "cite")`).

@@

## Aggregate outbreak surveillance counts (`r format(date_min_tallies, "%e %b %Y")`--`r format(date_max_tallies, "%e %b %Y")`)

On 24 March 2020, FSHCG introduced a bespoke reporting system to monitor Covid-19 cases in each care home for all care home residents and staff (including residents in contract beds). This new system requires care homes to report daily tallies of new cases identified among staff and residents to their FSHCG regional manager, and was used to report Covid-19 deaths to the Care Quality Commission (CQC).[@CQC2020stats] Daily counts were obtained for every care home:

- new symptomatic cases (residents)
- newly confirmed infection in home (residents)
- newly confirmed infection in hospital (residents)
- deaths related to Covid-19 (residents previously confirmed or with a death attributed to SARS-CoV-2 by coroner)
- symptomatic cases (staff)
- newly confirmed cases (staff)

In addition, the number of beds currently occupied was reported every Sunday. This weekly information was used to approximate daily occupancy using linear interpolation. Aggregate data were only extracted for residents in care homes for the elderly.

Care home characteristics were inferred from organisational data. Baseline care home occupancy was computed by averaging weekly occupancy before the first Covid-19 cases (between 1 January and 1 March 2020). This was used to calculate a ratio of the baseline occupancy to the number of bedroom ratio. We also estimated a ratio of the headcount of staff employed to the number of beds.

## Individual-level residents records (`r format(date_min, "%e %b %Y")`--`r format(date_max, "%e %b %Y")`)

Secondary use individual-level data for private residents receiving general and elderly care were extracted and pseudonymised by HW. They consisted of two datasets:

- **a residents dataset** containing one record per resident per care home, alongside residents' gender, date of birth, most recent dates of admission and discharge, date of first admission, type of stay (residential/nursing) and care (general, dementia, elderly). This dataset does not include any records for `r formatbm(sum(desc_homes[["Total contract beds"]], na.rm=T))` occupants of local authority 'block contract' beds.
- **an incidents dataset** containing `r formatbm(desc_before_exclusions$nrow_incidents)` reports filed by care home staff. Mandated report field were: resident forename, surname, care home identifier, incident date/time, and date/time of reporting. In addition, reports could record: date of birth, gender, information on Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death. Incidents were classified to indicate whether the resident was symptomatic, and whether an infection was confirmed on the basis of multiple variables (supplementary material 1).

HW created a resident index file bridging the two datasets using: the first three letters of the resident's forename; the resident's full surname; and the care home identifier. Clerical review of identifiers generated in this way established that no single identifier was allocated to distinct individuals in the residents' dataset. Just `r desc_before_exclusions$missing_id` (`r tbrounding(desc_before_exclusions$missing_id_pct)`%) incident records did not contain enough information to generate an identifier and were excluded. Identifiers of a further `r desc_before_exclusions$link_fail` records (`r desc_before_exclusions$link_fail_pct`%) could not be successfully linked to the residents dataset and were excluded, after clerical review established they could not be matched with a resident. A proportion of these records may have been filed in error, while others may relate to occupants of local authority beds who are not accounted for in the residents' dataset.

## Analysis

Incidence proportions and rates were calculated using the aggregate outbreak counts, as these were the trusted source of information used for reporting deaths to the CQC and encompassed all residents. The number of unique residents present in the homes between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")` was estimated by multiplying the number of private residents (from the individual-level dataset) by the ratio of the total resident-days (from the weekly occupancy censuses) to the total private resident-days (from the individual data). This approach assumes that the mean length of stay of local authority residents is identical to that of private residents (supplementary material 1). Aggregate outbreak counts were also used to estimate incidence and cumulative incidence across time using a life table approach. Daily occupancy was inferred from weekly occupied bed censuses using linear interpolation. As the total exposed-to-risk (susceptible resident-days) was unknown, it was approximated in a multiple decrements life table by making the assumption that previously symptomatic/confirmed residents were discharged at the same rate as susceptible residents (supplementary material 1). A Kaplan-Meier product limit estimator was applied to this life table to estimate the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths every day between `r format(date_min_tallies, "%e %B %Y")` and `r format(date_max_tallies, "%e %B %Y")`. The incidence proportion of confirmed cases in England was contrasted with incidence measured from a sample survey of English private households which tested nose/throat swabs from 26,995 individuals between `r format(england_results$date_min, "%e %B %Y")` and `r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]

Individual-level data (including dates of care home entry and exit) were available for private residents only. By assuming infection underreporting to affect residents conditionally at random, it was possible to use the data to study factors associated with the rate of infection and mortality. A Cox proportional hazards model was used to test the effect of both individual-level risk factors (sex, age, care type) and organisation-level risk factors (region, care home size, staff-to-bed ratio, occupant-to-bedroom ratio).

). We also compared care ratesDates of infectiondischarge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality in care homes withbased on the presence of symptoms and without outbreaks, defined as at least one /or confirmed case or Covid-19 related death during the study period. infection. A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (susceptible; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed). We also compared rates of infection and mortality in care homes with and without outbreaks, defined as at least one confirmed infection or Covid-19 related death during the study period. A time-dependent Cox proportional hazards model analysed the effect of the same factors on all-cause mortality based on the time-dependent infection classification (uninfected; symptomatic not confirmed; asymptomatic confirmed; symptomatic confirmed) as well as the time-dependent home outbreak status (no Covid-related death or confirmed infection; at least one Covid-related death or confirmed infection). The associated hazard ratios enable us to compare the mortality in each group as compared to the reference category (uninfected in non-outbreak care homes). The corresponding attributable fraction is obtained by using the model to predict the counterfactual mortality, then computing AFW, the attributable fraction within study:[@Samuelsen2008]

$$ AFW = \dfrac{\sum_{i=1}^{n}(\hat{F}_i - \hat{F}_i^\star)}{\sum_{i=1}^{n}\hat{F}_i} $$

where $\hat{F}_i$ designates the predicted probability of death of individual $i = \{1, \dots, n\}$ based on the observed infection/outbreak classification, and  $\hat{F}_i^\star$ the same probability had the individual been uninfected and living in a non-outbreak care home. Observed censoring times are used.

Bedroom occupancy and staffing ratios were estimated on the basis of baseline occupancy averaged before the first cases (between 1 January and 1 March 2020), so as to rule out reverse causality. Dates of discharge were used alongside the death indicator to estimate the case-fatality rate in home and hospital, as well as compare all-cause mortality based on the presence of symptoms and/or confirmed infection. Confirmed infection data in both sources can only be treated as underestimates of the true incidence, due to the gradual increase in nasopharyngeal swab (PCR) testing across the study period following the introduction of whole care home testing on 11 May 2020.[@dhscTesting2020]. By 22 June 2020, 90% of FSHCG care homes had participated in the whole home testing programme.

Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.

@@

The study population consisted of an estimated `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland. Two thirds (`r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes`) of homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, reported at least one confirmed infection or Covid-related death, either through Datix or outbreak surveillance counts.

>>>>

---
title: "Covid-19 infection and attributable mortality in residents of UK care homes: Cohort study using electronic records (March-June 2020)"
# author: "Dutey-Magni, PF, Williams, H, Hayward, A, Shallcross, L"
bibliography: references.bib
csl: sage-vancouver.csl
output:
  word_document:
    toc: yes
    reference_docx: template.docx
  html_document:
    df_print: paged
    toc: yes
    code_folding: hide
link-citations: yes
always_allow_html: yes
---

Peter Dutey Magni,$^{12}$ Haydn Williams,$^{3}$ Arnoupe Jhass,$^{4}$ Greta Rait,$^4$ Harry Hemingway,$^{125}$ Andrew Hayward,$^{6}$ Laura Shallcross$^{12*}$

$^1$ Institute of Health Informatics, University College London, London, UK

$^2$ Health Data Research UK, University College London, London, UK

$^3$ Four Seasons Health Care Group, Norcliffe House, Station Road, Wilmslow, Cheshire, UK

$^4$ Primary Care & Population Health, University College London, London, UK

$^5$ University College London Hospitals NIHR Biomedical Research Centre, London, UK

$^6$ Institute of Epidemiology & Health Care, University College London, London, UK

$*$ Corresponding author

@@

incidence_CT_residents <- table_nums(
  name = "incidence_CT_residents",
  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections in residents according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
supp_epi_CT_oubreak_only <- table_nums(
  name = "incidence_CT_staff",
  caption =  paste0("Cumulative incidence and rate of SARS-CoV-2 infections among staff according to FSHCG manager counts (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_resident_descriptives <- table_nums(
  name = "tab_resident_descriptives",
  caption = paste0("Characteristics of FSHCG private bed residents by care home type, sex, age, region and status on study exit (", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CFR_age_sex <- table_nums(
  name = "tab_CFR_age_sex",
  caption =  paste0("All-cause case-fatality rates by age and sex among private residents (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  "; ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_CoxPH_confirmed <- table_nums(
  name = "tab_CoxPH_confirmed",
  caption =  paste0("Risk factors for confirmed infection in private residents: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_overview$resident_id)),
  ")"))
tab_CoxPH_death <- table_nums(
  name = "tab_CoxPH_death",
  caption =  paste0("Risk factors for all-cause mortality in private residents of care homes with and without Covid-19 outbreaks: hazard ratios (HR) from a Cox proportional hazards model (n=", 
  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))
tab_attributable_deaths <- table_nums(
  name = "tab_attributable_deaths",
  caption =  paste0("Model-based estimates of attributable death in private residents of care homes with and without Covid-19 outbreaks (n=", 
  formatbm(dplyr::n_distinct(residents_episodes$resident_id)),
  ", ", format(date_min, "%e %b %Y"), "-", format(date_max, "%e %b %Y"), ")"))

@@

**Background** Rates of Covid-19 infection have declined in many countries, but outbreaks persist in care home residents who are at high risk of severe outcomes. Epidemiological data from care homes are scarce. We used population-level active surveillance to estimate incidence of, and risk factors for Covid-19, and attributable mortality in care home residents.

**Methods:** Cohort study using individual-level electronic health records from 8,713 residents and daily counts of infection for `r formatbm(desc_residents$CT_total_residents)` residents and `r formatbm(desc_staff$total_staff)` staff across `r dplyr::n_distinct(residents$home_code)` UK care homes. We modelled risk factors for infection and mortality using Cox proportional hazards and estimated attributable fractions.

**Findings:** `r formatbm(desc_residents$CT_total_symptomatic)` residents developed Covid-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory confirmed infections. Confirmed infection incidence in residents and staff respectively was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one Covid-19 infection or death. Lower staffing ratios and higher occupancy rates were independent risk factors for infection.

`r formatbm(desc_residents$DAT_total_deaths)` all-cause deaths occurred in `r formatbm(desc_residents$DAT_total_residents)` (`r tbrounding(desc_residents$DAT_p_death)`% [`r tbrounding(desc_residents$DAT_p_death_lower)`%; `r tbrounding(desc_residents$DAT_p_death_upper)`%]) residents. `r desc_residents$DAT_CFR$y` deaths occured in `r desc_residents$DAT_CFR$n` residents with confirmed infection (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Mortality in residents with no direct evidence of infection was 96% higher in care homes with at least one Covid-19 infection or death compared to care homes without (adjusted HR 1·9 [1·7; 2·2]). 33% of all-cause deaths were attributable to Covid-19 either through confirmed infection, or excess mortality.

**Interpretation:** 1 in 5 residents had symptoms of infection during the pandemic, but many cases were not tested. Higher occupancy and lower staffing levels increase infection risk. Disease control measures should integrate active surveillance and testing with fundamental changes in staffing and care home occupancy to protect staff and residents from infection.

**Funding:** Economic and Social Research Council [ES/V003887/1].

@@

Globally the number of cases of Covid-19 continues to increase, but cases in Europe have declined since April 2020,[@ECDC2020] following the introduction of lockdown measures.[@Hale2020] Although the incidence of infection in the general population in England is low (0.04%),[@cisbulletin2020] new infections persist, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] This raises the possibility that these settings represent a reservoir for transmission of infection back to the community.

In the UK, there are an estimated 400,000 residents living in approximately 11,000 homes for the elderly.[@LaingBuisson2019] Care home residents are particularly vulnerable to Covid-19 due to their advanced age and high prevalence of comorbidity,[@Zhou2020] and their frequent exposure to infection through close contact with staff members, other residents and contaminated surfaces in the care home setting. At the peak of the pandemic, the number of deaths in care home residents was three-fold higher than the equivalent period in in 2019.[@onsDR2020] Care home staff also have higher aged-standardised rates of Covid-19 related mortality compared to other occupations.[@onsDR2020occ] National statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved Covid-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to Covid-19 infection is important, to fully assess the impact of the pandemic on care homes.

The development of public health strategies to protect the public, residents and staff from Covid-19 requires knowledge of the burden of and risk factors for infection in care home residents and staff, linked to outcomes. However, there is no syndromic surveillance for infection in care homes in England, and widespread one-off testing for SARS-CoV-2 using reverse transcriptase polymerase chain reaction (RT-PCR) was not established for care home staff and residents until 11 May 2020.[@dhscTesting2020] Prior to this, testing was only available for care home residents or staff who were admitted to hospital, or as part of Public Health England’s (PHE) outbreak investigations which permitted a maximum of five tests per care home. Consequently national estimates of incidence and prevalence will substantially underestimate the burden of infection in care home residents and staff.

In the absence of cohort studies or active surveillance, outbreak investigations provide the most reliable estimates of the burden of infection and case-fatality.[@McMichael2020; @Dora2020] An estimated 44% of English care homes have had at least one outbreak, with a living systematic review[@Salcher-Konrad2020] reporting substantial variation in cumulative incidence of infection (0%-72%) and case fatality (0-34%) in care home residents. A major limitation of outbreak investigations is that follow-up is usually less than 30 days,[@Salcher-Konrad2020] and investigations are only conducted in settings in which outbreaks have been detected.  Understanding the proportion of care homes with undetected cases is crucial for policy decisions around the frequency of and justification for regular testing in care homes.

To our knowledge, there are no studies which have employed population-level active surveillance in care homes throughout this pandemic to measure outcomes of both suspected and confirmed infection in residents and staff. We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest independent provider of residential and nursing care, with the aim of identifying strategies to protect care home staff and residents from future waves of infection. Our objectives were to estimate incidence of and risk factors for infection and mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection. We also estimated mortality attributable to Covid-19.

@@

## Study population and setting

Staff or residents living and working in care homes for the elderly between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`, which were run by the Four Seasons Healthcare Group (FSHCG) were eligible for study inclusion. The FSHCG provides a combination of residential and nursing care (for residents with medical conditions), which is primarily state funded. Most residents are permanent, but a small proportion receive temporary (respite) care.

In 2020, there were 9,568 beds, representing 9% of all beds in England, Scotland and Northern Ireland (supplementary material 1). 90% of FSHCG care homes participated in the whole care home testing programme, implying that all staff and residents were tested for Covid-19 at least once between 11 May and 22 June 2020.

## Data sources

Electronic record datasets collected by the FSHCG are primarily used for billing and monitoring care quality, but have also been used in research.[@Smith2020]

*Individual-level data*

The care home chain collects electronic health record data on all their residents except those occupying beds that are 'block contracted' to the local authority (855 beds). The dataset includes: dates of care home entry and exit, sex, date of birth, type of stay (residential/nursing) and care (general, dementia, elderly).  Individual-level data on incidents including infections are also reported via 'Datix' which records the residents name, care home identifier, incident date/time, date of birth, sex, Covid-19 symptoms, tests and test results, resident current location (home/hospital), and death. Information in Datix was used categorise residents’ infection status into four groups:  (A) residents with no evidence of infection (not tested and/or no symptoms); (B) symptomatic residents (symptoms and not tested or tested negative); (C) asymptomatic residents with confirmed infection (no symptoms but tested positive); and (D) symptomatic residents with confirmed infection (symptoms and tested positive) (supplementary material 1). The term 'confirmed' denoted a positive PCR test. Datix was also used to differentiate in-hospital and in care home deaths, and to identify Covid-19 related deaths. Individual level data on residents was linked to Datix reports by creating an index file based on a combination of forename, surname and the care home identifier (supplementary material 2).  1492/1880 (79%) of Datix reports were successfully linked. Individual-level records were available between `r format(date_min, "%e %B")` and `r format(date_max, "%e %B %Y")`.

*Aggregate data*

On 24 March the care home chain introduced a new system to report Covid-19 cases and deaths. This required care home managers to report daily tallies in residents (new symptomatic cases, new confirmed infection in home, new confirmed infection in hospital, deaths related to Covid-19) and staff (symptomatic cases, new confirmed cases). Data on staff deaths were not extracted due to the small number of cases. Covid-19 related deaths were defined as death in a resident with confirmed infection or a death attributed to Covid-19 by the coroner.  The number of occupied beds in each care home was reported weekly.

Care home characteristics (number of beds, region, nursing versus residential care) were collected from organisational data. We therefore extracted organisational data, individual-level data for 8713 residents and aggregate data for all care home staff and residents (`r fig_nums("fig_study_diagram", display = "cite")`).

@@

## Risk factors for infection

Risk factors for infection included individual-level variables (age, sex, general or dementia care, residential versus nursing care) and care home characteristics (number of beds, care home occupancy and staff to bed ratio). Baseline care home occupancy was computed by averaging weekly occupancy in January-March 2020, before the first Covid-19 case, in order to calculate a ratio of baseline occupancy to the number of bedrooms. We also estimated the ratio of staff to beds. A care home outbreak was defined as at least one confirmed infection or Covid-19 related death.

## Statistical analysis

*Infection in care home staff and residents*

Prevalence, incidence and cumulative incidence were calculated for residents and staff using the aggregate daily tallies. These were the trusted source of information used for national reporting of cases, and encompassed all residents and staff.[@CQC2020stats] Infection incidence was also estimated from Datix, but was subject to under-reporting (supplementary material 2). In order to calculate infection and death incidence, we estimated the total number of residents in each care home by extrapolating estimates of care home occupancy from the individual-level dataset (because dates of care home entry and exit were not available for 855 beds, supplementary material 1). Daily occupancy was inferred from the weekly report of bed occupancy using linear interpolation. The total number of residents at risk of infection was unknown, so it was approximated in a multiple decrements life table (supplementary material 1). The life table allowed us to compute Kaplan-Meier product limit estimators of the cumulative incidence of symptoms, confirmed infections, and Covid-19 related deaths by day. The rate ratio for care home versus community infections was estimated by contrasting the cumulative incidence for confirmed cases in England with estimates from a national household survey for the period `r format(england_results$date_min, "%e %B")`-`r format(england_results$date_max, "%e %B %Y")`.[@cisbulletin2020; @cisdata2020]

*Mortality, attributable mortality and risk factors*

Individual-level data were used to estimate rates of infection, all-cause mortality and case-fatality by age and gender in residents. Aggregate data were also used to estimate the crude rate of Covid-19 related mortality. Cox proportional hazards models were used to test the association between individual and organisational-level risk factors and confirmed infection. In order to investigate the relationship between Covid-19 infection and excess mortality, we assumed that residents in non-outbreak care homes had not been exposed to infection, and would therefore not experience excess Covid-19 related mortality.  We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.

A Cox proportional hazards regression model tested the effect of individual and care home level risk factors on all-cause mortality, alongside the effect of the time-variant infection status (groups A-D) and care home outbreak status.  We estimated the attributable fraction of deaths for each infection category in care homes with and without outbreaks, taking the reference category as individuals with no direct evidence of infection (group A) in non-outbreak care homes. This fraction was obtained by using the model to predict the counterfactual mortality, then computing the attributable fraction within study.[@Samuelsen2008] Ninety-five percent confidence intervals for proportions and rates were computed from the exact Poisson and binomial limits.

Data were analysed in `R`3·5·0 using the `epitool`[@Aragon2020] and `survival`[@therneau2015] libraries. All computer syntax is available on an online repository.[@Dutey2020]

@@

The study included `r formatbm(desc_residents$CT_total_residents)` residents across England, Scotland and Northern Ireland and `r formatbm(desc_staff$total_staff)` staff. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes, totalling `r formatbm(desc_residents_outbreak_only$CT_total_residents)` residents, recorded at least one Covid-19 outbreak (including unconfirmed outbreaks) in either the individual-level or aggregate datasets.  The mean duration of follow-up for residents and staff was `r round(desc_residents$CT_total_cdeath_denom/desc_residents$CT_total_residents)` days and `r as.integer(date_max_tallies-date_min_tallies)` days respectively in the aggregate dataset. Mean and median duration of resident follow-up was `r round(mean(residents_overview$resident_days))` and `r median(residents_overview$resident_days)` days respectively in the individual-level dataset.
